Higher-Order unification for free!

### 2024-04-22 paper/main.tex
minor rewordings

The second step involves allocating new variables in the
memory of \Fo{}. This technicality is required because some
higher-order unifications may require pruning a variable. For example,
$F\appsep x\appsep y = F\appsep x\appsep z$ requires allocating a 
variable $G$ in order to express the assignment $F_{ab} \mapsto G_a$.

The final step is to decompile each assignment.
Decompiling a term is straightforward since \mapStore is a bijection.
The only complex part concerns the \elpiIn{abs} node. In our simple setting,
the \elpiIn{flam} node carries no additional information (other than the function body),
so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
However, in the case of Coq, where lambdas carry the type of the bound variable, one
must store this information somewhere. Note that this information is similar
to the arity of variables; that in Coq, unification variables have a
(function) type, and this type can be used to annotate the lambdas
needed to express their assignment.

@@

\noindent
The statement above roughly states that the translation from
\Fo{} to \Ho is not lossy, it can be reversed. The statement
is made convoluted by the fact that we hid the procedure to
decompile a term in favor of the one to decompile substitutions.
We omit to sketch the proof of this property for brevity.

@@

\noindent
So far the compiler is very basic. It does not really
enforce that the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first-order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will enforce \cref{prop:w-enforcing}
the proof will be adjusted to cover for the new cases.

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification (as per \cref{prop:good-unif})
  for \Eo in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb{(\foUnifPb)}$.
\end{theorem}
\begin{proof}[Proof sketch]
Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same,
and by \cref{prop:uniffow} \Ue is equivalent to \Uo.
\end{proof}

@@

In the following sections, we adopt this notation discuss the compiler's
output: \foUnifPb represents the input set of problems that compiled to
\hoUnifPb with memory mapping \mapStore and links \linkStore. For example:
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
  {{{\eta,\Gamma,a,b}}}

We index each sub-problem, sub-mapping, and sub-link with its position
starting from $1$ and counting from left to right, top to bottom.
For example, $\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
%The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.

@@

In order to make \Uo higher-order, we need to
take care of terms in \maybebeta.
In the example below, we can see that the basic compilation given in 
the previous section is not able to make the \Ho unification problem succeeds.
%
% make test ONLY=7006 TEX=tex
\printAlllSingle
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}
The unification problem $\hoUnifPb_1$ fails while trying to unify
$A \appsep x$ and $x$, which is equivalent to <<\elpiIn{app [uva A [], x]}>>
versus \elpiIn{x}.
In order to exploit the higher-order unification algorithm of the meta language,
we %need to 
compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$.
%,which is \elpiIn{uva A [x]}.

@@

\paragraph{Decompilation} Since no link is created by the compilation of 
\maybebeta terms, no modification to the \elpiIn{commit-link}
is needed.

\paragraph{Progress} Similarly to decompilation, since no link is produced,
no modification to the \elpiIn{progress} predicate is needed.

@@

\section{Handling of \maybeeta}\label{sec:eta}
A term 
$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
$x$ does not occur free in $t$, and conversely, $t$ is the $\eta$-contraction of
$\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
but the current compilation scheme does not,
as shown by the following example:
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

\noindent
While $\lambda x.X \appsep x \Uo{} f$ does admit the solution
$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
\hoUnifPb does not:
\elpiIn{lam x\ uva A [x]} and
\elpiIn{con"f"} start with different, rigid, term constructors hence
\Ue{} fails.

>>>>

The second step allocates
in the memory of \Fo{}
new variables used to implement the pruning of higher-order variables. 
For example,
$F\appsep x\appsep y = F\appsep x\appsep z$ requires allocating a 
variable $G$ in order to express the assignment $F_{ab} \mapsto G_a$.

The final step is to decompile each assignment. The \elpiIn{val}
node carries a term that easy to decompile since \mapStore is a bijection.
Since the \elpiIn{flam} node
carries no additional information (other than the function body),
each \elpiIn{abs} node can be decompiled to a \elpiIn{flam} one.

However, if the \elpiIn{flam} node was carrying more information,
as is the case for Coq where it holds the type of the bound variable,
one would need to store this piece of information in the memory map,
were we store the arity (that is a very simple function type).

@@

% \noindent
% The statement above roughly states that the translation from
% \Fo{} to \Ho is not lossy, it can be reversed. The statement
% is made convoluted by the fact that we hid the procedure to
% decompile a term in favor of the one to decompile substitutions.
We omit to sketch the proof of this property for brevity.

@@

\noindent
So far the compiler is very basic. It does not really
enforce that the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first-order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will unsure \fullref{prop:w-enforcing}
and
the proof will be refined to cover for the new cases.

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb{(\foUnifPb)}$. \label{thm:fidfow}
\end{theorem}
\begin{proof}[Proof sketch]
Since \elpiIn{progress1} is a no-op then \fstep and \hstep are the same.
By \cref{prop:uniffow}, \Ue is equivalent to \Uo in \wellb.
\end{proof}

@@

In the following sections we use the following notation for input and output
of the compiler. \foUnifPb represents the input unification problems in \Fo{}
while \hoUnifPb their corresponding compiled version with memory mapping
\mapStore and links \linkStore. For example:
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
  {{{\eta,\Gamma,a,b}}}

  \noindent
We index each sub-problem, sub-mapping, and sub-link with its position
starting from $1$ and counting from left to right, top to bottom.
For example, $\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.

As we deal with each category of problematic terms we
weaken the assumptions of~\cref{thm:fidfow} using the following
definitions:

%The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.

@@

% In order to make \Uo higher-order, we need to
% take care of terms in \maybebeta.
% In the example below, we can see that
The basic compiler given in 
the previous section is unable to make the following
higher-order unification problem succeed.
%
% make test ONLY=7006 TEX=tex
\printAlllSingle
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}
The unification problem $\hoUnifPb_1$ fails while trying to unify
$A \appsep x$ and $x$, which is equivalent to <<\elpiIn{app [uva A [], x]}>>
versus \elpiIn{x}.
In order to exploit the higher-order unification algorithm of the meta language,
we %need to 
compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$.
%,which is \elpiIn{uva A [x]}.

@@

\paragraph{Decompilation}
No change
to \elpiIn{commit-link}
since no link is added.
% by the compilation of 
%\maybebeta terms, no modification to the \elpiIn{commit-link}
%is needed.

\paragraph{Progress}
No change to \elpiIn{progress} since no link is added.
% Similarly to decompilation, since no link is produced,
% no modification to the \elpiIn{progress} predicate is needed.

@@

\section{Handling of \maybeeta}\label{sec:eta}
A term 
$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
$x$ does not occur free in $t$, and conversely, $t$ is the $\eta$-contraction of
$\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
but the current compilation scheme does not,
as shown by the following example:
while $\lambda x.X \appsep x \Uo{} f$ does admit the solution
$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
\hoUnifPb does not.
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

\noindent
The reason is that \elpiIn{lam x\ uva A [x]} and
\elpiIn{con"f"} start with different, rigid, term constructors.
